---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---



## Loading and preprocessing the data
###Libraries
These are the libraries we are going to use:
```{r, results="hide", message=FALSE, warning=FALSE}
library(dplyr)
library(lubridate)
library(ggplot2)
```

### Loading the data
We begin by loading the activity data CSV file into a variable called data.
```{r}
data <- read.csv("activity.csv", na.strings="NA")
```
### Transforming the data
Let's look to what we've loaded
```{r}
head(data)
```  
Date and interval types:
```{r}
class(data$date[0])
class(data$interval[0])
```
It is useful to merge these columns into a single, date-class column. Let us call this new column DateFull:

```{r}
data <- mutate(data, dateFull = ymd_hm(paste(date, sprintf("%04d", interval))))
head(data)
class(data$dateFull[0])
```

Now, we are ready to to operate with date functions.


## What is mean total number of steps taken per day?
First, we group the data by date:
```{r}
byday <- group_by(data, date)
```
Because date has a resolution of day, we can be sure that byday holds a correct day-by grouping.

### Steps per day
The total steps per day is calculated by summarizing byday with sum:
```{r}
steps_per_day<-summarise(byday, steps=sum(steps) )
steps_per_day
```

### A histogran of steps per day
```{r}
hist(steps_per_day$steps, main="Historgram of steps per day", xlab="steps")
```

### Mean and median of the steps taken per day
A call to summary gives:
```{r}
summary(steps_per_day$steps)
```

## What is the average daily activity pattern?
First, we  group data by interval, then summarize with mean:
```{r}
byinterval <- group_by(data, interval)
steps_by_interval <- summarise(byinterval, steps=mean(steps, na.rm=TRUE))
```
### Average steps vs interval (across all days)
```{r}
plot(steps_by_interval, type='l')
```
.

### Which interval holds the maximum ?
To know which 5-minute inteval contains the maximum number of steps, we use which.max over our summarized data to get the index:
```{r}
max_ind <- which.max(steps_by_interval$steps)
steps_by_interval[max_ind,]
```
We can see then, that the 8:35AM time its the one when more average steps occurr, with a value of 206.



## Imputing missing values

### How many missing values we have
```{r}
sum(is.na(data$steps))
mean(is.na(data$steps))
```
2304 missing values, which is the 13% of the data set.

### Filling NAs
Let us try to fill NAs with the hourly mean. First, we create a new dateHourly column, constructed by removing the minutes and seconds from the dateFull col.
```{r}
tmpdata <- mutate(data, dateHourly=ymd_h(as.character(dateFull, "%Y%m%d %H") ))
```
Now, lets create a data frame with the steps averaged by the hour.
```{r}
byhour <- group_by(tmpdata, dateHourly)
step_hour_means <- summarise(byhour, steps_mean = mean(steps) )
```
We merge the two dataframes, and add a new column called steps_filled. This column will hold the original step value if not NA, or will hold the hour averege instead
```{r}
tmpdata <- merge(tmpdata, step_hour_means)
tmpdata <- mutate(tmpdata, steps_filled=ifelse(is.na(steps), steps_mean, steps))
```
#### Did it work ?
Summary for steps_filled
```{r}
summary(tmpdata$steps_filled)
```
Summary for steps
```{r}
summary(data$steps)
```
This gives us the same number of missing NAs. 

#### Why dit it fail?
By examining the data we can see that an hour-based mean is not useful becasue missing values are present in  whole-days blocks, meaning whole days are missing. Now, becasue we don't have any data for such a day, there doesn't seem to exist a good mean-injection value we can use to fill NAs.

### New data set without NAs
A better approach will be then to simple eliminate rows with missing values:

```{r}
fdata <- data[complete.cases(data),]
summary(fdata$steps)
```
Which gives us the same summary as before, but without NAs. 

Our new working dataset will be fdata.

### Analyzing the new set

#### Let us do a histogram.
First, we  group data by interval, then summarize with mean.
```{r}
byinterval <- group_by(fdata, interval)
steps_by_interval <- summarise(byinterval, steps=mean(steps, na.rm=TRUE))
plot(steps_by_interval, type='l')
```
.

#### Mean and median:
From our new NA-removed data
```{r}
summary(fdata$steps)
```
From our original data
```{r}
summary(data$steps)
```

#### In conclusion
We got the same histogram, mean and mediam as when we used the full data set.

By simply removing missing data, we are obtaining the same results.




## Are there differences in activity patterns between weekdays and weekends?

### New factor, day_type
First, we create a new column called wday, which holds the week day's name, and then, a factor called day_type with either  'weekend' or 'weekday' values.
```{r}
fdata <- mutate(fdata, wday=weekdays(dateFull, abbreviate=TRUE))
fdata <- mutate(fdata, day_type=ifelse(wday=='Sat' | wday=='Sun', 'weekend', 'weekday'))

```
### Weekend vs weekday plot
To get an interval steps, splitted by day_type, we group by interval and day_type, then, summarize with mean:
```{r}
grouped <- group_by(fdata, interval, day_type)
grouped_mean <- summarise(grouped, steps=mean(steps))
```

Using qplot we can have two plots of  steps vs interval one for each day_type:
```{r}
qplot(data=grouped_mean, x=interval, y=steps, facets=day_type~. , geom="line", color=day_type)
```

### Analysis
This plot shows two differences in activity patterns:

1. Weekends' activities start later.
2. Weekends' activity are more evenly distributed trhu the day, whereas weekdays' mornings are clearly peaking.


#### end.

